Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2025-01-21 LICS/sections/e-hypergraphs.tex
wip

Hierarchical (hyper-)graphs are a well-studied area of research \cite{plump:hierarchical-graphs, montanari:gs-lambda, palacz:hierarchical-transform, Gaducci:hierarchical-graphs, Ghica:hierarchical}.  
In this section,  we follow analogous steps to those used to achieve a combinatorial representation of $\catname{S}(\Sigma)$ by $\mathbf{MHypI}(\Sigma)$,  but in the semilattice-enriched setting.  
First we define an \textit{e-hypergraph},  which supports a hierarchical notion of ``e-box'' with distinct components. 
Then,  we extend this notion with interfaces via a generalisation of the cospan construction.  
Finally,  we restrict the morphisms under consideration to those satisfying the MDA condition,  as in the previous section.  
An example diagram representing e-hypergraphs with interfaces can be found in Figure \ref{fig:A+B}.

@@

\begin{definition}[E-hypergraphs]
\label{def:e-homo}    
An \emph{e-hypergraph $\mathcal{G}$ over a signature $\Sigma$} is a tuple $(V,E,s,t,l,<,\consistency)$,  where $(V,E,s,t)$ is an unlabelled hypergraph;  $l :E \to \Sigma + 1$ is a labelling function modified to include an extra value $\bot$ in its codomain; $<$ is a strict partial order on $V + E$ called the \textit{child relation}, for which we introduce the following additional notation.
For $x \in V + E$,  the set of \emph{parents} of $x$ is denoted as $[x) = \{x' ~|~ x' < x\; \}$.
% \[
% 	[x) = \{x' ~|~ x' <_c x\; \} \qquad (x] = \{x' ~|~  x <_{c} x'\}
% \]
We will denote the immediate parent $e$ of $x$ by writing $e <^{\mu} x$.
We call edges $e$ such that $l(e) = \bot$ \textit{hierarchical},  and edges $e$ and vertices $v$ such that $[e) = \varnothing$ and $[v) = \varnothing$ \emph{top-level}.
An edge $e$ is called maximal if $\{x' | e < x'\} = \varnothing$.
We require that the child relation satisfies the following conditions:
\begin{enumerate}
	\item each parent set contains exclusively hierarchical edges;
	\item each $x$ has at most one immediate parent;
	\item the maximal edges must be labelled;
	\item if $v \in s(e)$ then $e' <^{\mu} e$ iff $e' <^{\mu} v$ and similarly if $v \in t(e)$.
\end{enumerate}
Finally, $\consistency$ is a \textit{consistency relation} which is given by the union of a family of equivalence relations $\consistency_p$ on each set $\{x \in V + E ~|~ p <^\mu x\}$ of elements which share the same parent where each relation is also closed under connectivity, \textit{i.e.}, if $v \in s(e)$ or $v \in t(e)$ such that $p <^{\mu}(v)$ and $p <^{\mu}(e)$ then $v \consistency_{p} e$.
We require that $\consistency_{p} \not = (E_{p} + V_{p}) \times (E_{p} + V_{p})$ where $V_{p} = \{ v ~ | ~ p <^{\mu} v\}$ and $E_{p} = \{ e ~ | ~ p <^{\mu} e\}$.
% We collect the union of all these equivalence relations into one relation $\consistency = \bigcup_{p} \consistency_p$,  which is defined on a subset of $V+E$. 
% Note that this relation is symmetric and transitive if considered on $V + E$.
% The consistency relation must additionally satisfy that if $v \in s(e)$ or $v \in t(e)$ and there exists $e'$ such that $e' <^{\mu} v$ and $e' <^{\mu} e$ then $v \consistency e$.
\label{def:consistency_properties}
\end{definition}
Intuitively,  the child relation equips a hypergraph with a hierarchical structure: certain edges,  labelled with $\bot$,  are meant to interpret the dashed e-boxes of string diagrams.  
These edges can be designated as "parents" to sub-e-hypergraphs,  which interpret the various components of the e-boxes.  
These components are distinguished by the consistency relation, which partitions the maximal sub-e-hypergraph below the hierarchical edge.
% Technically,  condition $(4)$ on the child relations says that $<^{\mu}$ respects connectivity of edges and vertices,  and similarly for the consistency relation. 
An example of an e-hypergraph can be seen below.
\[
\adjustbox{scale=0.75}{
\tikzfig{../figures/combinatorial_semantics/e-hypergraph-example-2}
}
\]

This e-hypergraph corresponds to the following tuple
\[
(V = (v_1, v_2, v_3, v_4, v_5, v_6\}, E = \{e_1, e_2, e_3\}, s, t, l, <, \consistency)
\]
\begin{align*}
s(e_1) = \{v_1\} &\; t(e_1) = \{v_2\} &\; l(e_1) = \bot \\
s(e_2) = \{v_3\} &\; t(e_2) = \{v_4\} &\; l(e_2) = e_2 \\
s(e_3) = \{v_5\} &\; t(e_3) = \{v_6\} &\; l(e_3) = e_3 \\
\end{align*}

and $[v_3) = [v_4) = [v_5) = [v_6) = [e_2) = [e_3) = \{e_1\}$ and $[v_1) = [v_2) = [e_1) = \varnothing$.
$v_3 \consistency v_4 \consistency e_2$ and $v_5 \consistency e_3 \consistency v_6$ and the relation is undefined for the rest vertices and edges.
Because $v_3 \not \consistency v_5$ and because the relation is closed under connectivity we delimit the connected components by a dashed vertical line.
The condition $\consistency_{p} \not = (E_{p} + V_{p}) \times (E_{p} + V_{p})$, in particular, does not allow hierarchical edges with no delimiting lines.
$e_2$ and $e_3$ are maximal edges.
\begin{definition}[E-hypergraph homomorphism]
An \textit{e-hypergraph homomorphism} $\phi: \mathcal{F} \to \mathcal{G}$ of e-hypergraphs $\mathcal{F},\mathcal{G}$ is a pair of functions $\phi_V : V_{\mathcal{F}} \to V_{\mathcal{G}}, \phi_E : E_{\mathcal{F}} \to E_{\mathcal{G}}$ such that
\begin{enumerate}
    \item $\phi$ is a hypergraph homomorphism
    \item for $v \in V_{\mathcal{F}}$ and $e \in E_{\mathcal{F}}$
	    \[
        \text{if } e <_{\mathcal{F}}^{\mu} v \text{ then } \phi(e) <_{\mathcal{G}}^{\mu} \phi(v)
        \]
        and for $e_1, e_2 \in E_{\mathcal{F}}$
        \[
        \text{if } e_1 <_{\mathcal{F}}^{\mu} e_2 \text{ then } \phi(e_1) <_{\mathcal{G}}^{\mu} \phi(e_2)
        \]
        \item for all $x_1, x_2 \in \mathcal{F}$,  $x_1 ~\consistency_{\mathcal{F}}~ x_2$ implies $\phi(x_1) ~\consistency_{\mathcal{G}}~ \phi(x_2)$. 
\end{enumerate}
% If we further have that if $[x_1) = \varnothing$ then $[\phi(x_1)) = \varnothing$ we call the homomorphism \textit{strict}.
\end{definition}
The conditions on e-hypergraph homomorphisms require to preserve immediate parents and the consistency relation.  
\begin{definition}[Category of e-hypergraphs]
The \emph{category of e-hypergraphs},  denoted  $\catname{EHyp(\Sigma)}$,  has e-hypergraphs as objects and e-hypergraph homomorphisms as morphisms.  
\end{definition}

The category of e-hypergraphs has coproducts,  given by the disjoint union of e-hypergraphs,  and an initial object given by the empty e-hypergraph.  
A concrete description of the pushout of two morphisms in this category is given in the Appendix.  
Generally,  the pushout of two e-hypergraph homomorphisms need not exist,  but we prove that it does when certain technical conditions are satisfied, including the requirement that the shared domain of the morphisms that form the span for the pushout is a discrete e-hypergraph.

@@

We will call $n$  \textit{external input interfaces} and $n'$ \textit{internal input interfaces}.
We call $n' \setminus f_{ext}(n)$  the \textit{strictly internal input interfaces}.  
We do analogously for the \emph{output interfaces},  with respect to $m$,  $m'$ and $m' \setminus g_{ext}(m)$.  
We occasionally conflate $f_{ext}$ with $f_{ext};f_{int}$ when it is clear from context,  and also conflate $n$ and $m$ with their images in $n'$ and $m'$,  and likewise $m, m, n', m'$ with their images in $\mathcal{G}$. 
Given an edge $e \in E_\mathcal{G}$ such that $l(e) = \bot$,  we call the \textit{inputs of $e$} the intersection of the strictly internal input interface of $\mathcal{G}$ with the immediate successors of $e$,  and analogously for the \emph{outputs of $e$}.

Consider a relation $R = \{x R y \text{ if } f_{int}(x) \consistency f_{int}(y) \text{ for } x,y \in n'\}$ and let $S$ be its reflexive closure.
The latter partitions $n'$ into non-empty subsets $\{p_{j}\}_{j=1}^{k}$. We get an analogous partition for $m'$.
An example of an extended cospan can be seen in Figure~\ref{fig:e-cospan-example}.
The cospan represents a morphism that encodes symmetry put into a hierarchical edge.
The partition of the interfaces is given with colours.
The labels suggest how $f_{int}, f_{ext}$ (respectively, $g_{int}, g_{ext}$) act on vertices.
The ordering of vertices in the interfaces is from left to right.

@@

\begin{definition}
\label{def:iso}
Two extended cospans are isomorphic if there exist isomorphisms $\alpha$, $\beta$ and $\gamma$ making the following diagram commute and such that $\alpha$ and $\gamma$ preserve order within each $p_i$.
\[
\scalebox{0.75}{
    \tikzfig{../figures/combinatorial_semantics/isomorphic_e_cospans}
}
\]
\end{definition}
Intuitively, such construction suggests that the position of the image of $n$ within $n'$ relative to the strict internal interface does not matter; and that when considering individual components of hierarchical edges with their respective portions of strict internal input and output interfaces as plain cospans of hypergraphs (by only considering top-level vertices and edges of these components) they should be isomorphic cospans of hypergraphs in the sense from~\cite{bonchi_string_2022-1}.
Examples of isomorphic and non-isomorphic extended cospans can be found in Appendix.

@@

Composition of two morphisms 
\begin{align*}
	n \xrightarrow{f_{ext}} n' \xrightarrow{f_{int}} &\mathcal{F} \xleftarrow{f'_{int}} m' \xleftarrow{f'_{ext}} m\\
	m \xrightarrow{g_{ext}} m'\!' \xrightarrow{g_{int}} &\mathcal{G} \xleftarrow{g'_{int}} k' \xleftarrow{g'_{ext}} k
\end{align*} is computed in two stages.
First, $\mathcal{H}$ is computed as the result of the pushout square shown below; 
% https://q.uiver.app/#q=WzAsMTAsWzAsMCwiZXh0KFgpIl0sWzEsMSwiWCJdLFsyLDIsIkEiXSxbMywwLCJZIl0sWzQsMCwiZXh0KFkpIl0sWzUsMCwiWSciXSxbNiwyLCJCIl0sWzcsMSwiWiJdLFs4LDAsImV4dChaKSJdLFs0LDMsIl57XFx1bGNvcm5lcn1BK197Zl97ZXh0fTtmLGdfe2V4dH07Z31CIl0sWzAsMV0sWzEsMl0sWzMsMiwiZiJdLFs0LDMsImZfe2V4dH0iXSxbNCw1LCJnX3tleHR9IiwyXSxbNSw2LCJnIiwyXSxbNyw2XSxbOCw3XSxbMiw5XSxbNiw5XV0=
\[
\trimbox{0cm 0cm 0cm 0.75cm}{
\adjustbox{scale=0.9,center}
{\tikzfig{../figures/combinatorial_semantics/pushout_e_cospans}}
}
\]
then,  the result of composition is defined as follows:
\[
n \xrightarrow{f_{ext};\iota_1} n' + (m'' \setminus m) \xrightarrow{h_{1}} \mathcal{H} \xleftarrow{h_2} k' + (m' \setminus m) \xleftarrow{g'_{ext};\iota_1} k
\]
where $h_i$ are defined as follows,  using $|$ to denote the restriction of a function on a discrete e-hypergraph. 
\[
    h_1 = [ f_{int};p_1, ~(g_{int};p_2)|_{m'' \setminus m} ]
\qquad
    h_2 = [ g'_{int};p_2, ~(f'_{int};p_1)|_{m' \setminus m} ]
\]
% \[
% \begin{tikzcd}
% Y_{ext} \arrow[d] & 0 \arrow[l] \arrow[d]    \\
% Y'          & Y' \setminus Y_{ext} \arrow[l]
% \end{tikzcd}
% \]
% and then the disjoint union is just a coproduct. 
% where $g'$ is the restriction of $g$ to ${Y' \setminus Y_{ext}}$ and $f'$ the restriction of $f$ to $Y \setminus Y_{ext}$. 
The identity of composition is given by the obvious extended cospan of identities. 
$\Ecospans$ inherits a symmetric monoidal structure from the coproduct (and initial object) structure of $\catname{EHyp({\Sigma}})$,  analogously to Definition~\ref{def:cspd}.  
\end{definition}

@@

\begin{definition}[Category of well-typed MDA e-hypergraphs with interfaces]
We call an extended cospan in $\Ecospans$,  as below,   \textit{monogamous directed acyclic (MDA)} if
\[
n \xrightarrow{f_{ext}} n' \xrightarrow{f_{int}} \mathcal{G} \xleftarrow{g_{int}} m' \xleftarrow{g_{ext}} m
\]
\begin{enumerate}
        \item underlying hypergraph of $\mathcal{G}$ is directed acyclic; 
        \item $f_{int}$, $g_{int}$ are monomorphisms; 
        \item The in-degree and out-degree of every vertex is at most 1; 
        \item Vertices with in-degree $0$ are precisely the image of $f_{int}$; 
        \item Vertices with out-degree $0$ are precisely the image of $g_{int}$; 
        % the item below is not assumed by the definition of a cospan of e-hypergraphs
        % \item $[v) = \varnothing$ for all $v$ in the image of $f_{ext};f$, and for all $v$ in the image of $g_{ext};g$. 
\end{enumerate}
\label{def:monogamy_ehyp}

Further,  consider the sets of input and output vertices,  $I$ and $O$ respectively,  of a hierarchical edge $e$ of an e-hypergraph $\mathcal{G}$. 
The consistency relation of $\mathcal{G}$ partitions $I$ and $O$.  
We call $e$ \emph{well-typed} if the size of each subset in the partition of $I$ is $|s(e)|$ and the size of each partition of $O$ is $|t(e)|$.
We will call an MDA extended-cospan in $\MdaEcospans$ \emph{well-typed} if all hierarchical edges in its apex are well-typed.

We denote by $\MdaEcospans$ the subcategory of $\Ecospans$ consisting of well-typed MDA extended-cospans. 
\end{definition}
Note,  $\MdaEcospans$ is indeed a category,  and in fact inherits the symmetric monoidal structure of $\Ecospans$:  it is easy to check that identities,  the monoidal unit,  symmetry,  and the composition and tensor product of any two well-typed cospans are all (again) well-typed.

Unlike in the previous section,  in order to give an equivalence with $\textbf{PROP}^+(\Sigma)$,  we must first develop the theory of DPOI rewriting for e-hypergraphs.  
This is because the equations for semilattice enrichment are not intended to be subsumed by the e-hypergraph representation, apart from the commutativity equation which is absorbed by the definition of isomorphic cospans.  
Instead,  we will treat them via rewriting,  as we do with equations of the signature.  This is because the semilattice equations involve sharing and copying, operations on string diagrams which are not usually quotiented away.
% , cf. string diagrams for Cartesian categories.  
An ideal combinatorial representation would factor out the two semilattice equations which \textit{do not} involve sharing and copying (namely,  associativity and idempotence) --- however, a limitation of the current work is that we do not achieve this.  
Presently,  we will treat these equations via rewriting as well.

>>>>

Hierarchical (hyper-)graphs are a well-studied area of research \cite{plump:hierarchical-graphs, montanari:gs-lambda, palacz:hierarchical-transform, Gaducci:hierarchical-graphs, Ghica:hierarchical}.  
In this section,  we follow analogous steps to those used to achieve a combinatorial representation of $\catname{S}(\Sigma)$ by $\mathbf{MHypI}(\Sigma)$,  but in the semilattice-enriched setting.  
First we define an \textit{e-hypergraph},  which supports a hierarchical notion of ``e-box'' with distinct components. 
Then,  we extend this notion with interfaces via a generalisation of the cospan construction.  
Finally,  we restrict the morphisms under consideration to those satisfying the MDA condition,  as in the previous section.  
% An example diagram representing e-hypergraphs with interfaces can be found in Figure \ref{fig:A+B}.

@@

\begin{definition}[E-hypergraphs]
\label{def:e-homo}    
An \emph{e-hypergraph $\mathcal{G}$ over a signature $\Sigma$} is a tuple $(V,E,s,t,l,<,\consistency)$,  where $(V,E,s,t)$ is an unlabelled hypergraph;  $l :E \to \Sigma + 1$ is a labelling function modified to include an extra value $\bot$ in its codomain; $<$ is a strict partial order on $V + E$ called the \textit{child relation}, for which we introduce the following additional notation.
For $x \in V + E$,  the set of \emph{parents} of $x$ is denoted as $[x) = \{x' ~|~ x' < x\; \}$.
% \[
% 	[x) = \{x' ~|~ x' <_c x\; \} \qquad (x] = \{x' ~|~  x <_{c} x'\}
% \]
We will denote the immediate parent $e$ of $x$ by writing $e <^{\mu} x$.
We call edges $e$ such that $l(e) = \bot$ \textit{hierarchical},  and edges $e$ and vertices $v$ such that $[e) = \varnothing$ and $[v) = \varnothing$ \emph{top-level}.
An edge $e$ is called maximal if $\{x' | e < x'\} = \varnothing$.
We require that the child relation satisfies the following conditions:
\begin{enumerate}
	\item each parent set contains exclusively hierarchical edges;
	\item each $x$ has at most one immediate parent;
	\item the maximal edges must be labelled;
	\item if $v \in s(e)$ then $e' <^{\mu} e$ iff $e' <^{\mu} v$ and similarly if $v \in t(e)$.
\end{enumerate}
Finally, $\consistency$ is a \textit{consistency relation} which is given by the union of a family of equivalence relations $\consistency_p$ on each set $\{x \in V + E ~|~ p <^\mu x\}$ of elements which share the same parent where each relation is also closed under connectivity, \textit{i.e.}, if $v \in s(e)$ or $v \in t(e)$ such that $p <^{\mu}(v)$ and $p <^{\mu}(e)$ then $v \consistency_{p} e$.
We require that $\consistency_{p} \not = (E_{p} + V_{p}) \times (E_{p} + V_{p})$ where $V_{p} = \{ v ~ | ~ p <^{\mu} v\}$ and $E_{p} = \{ e ~ | ~ p <^{\mu} e\}$.
Given an e-hypergraph $\mathcal{G}$ we can consider a corresponding \textit{underlying} hypergraph $\mathcal{G}'$ by forgetting $<$ and $\consistency$.
% We collect the union of all these equivalence relations into one relation $\consistency = \bigcup_{p} \consistency_p$,  which is defined on a subset of $V+E$. 
% Note that this relation is symmetric and transitive if considered on $V + E$.
% The consistency relation must additionally satisfy that if $v \in s(e)$ or $v \in t(e)$ and there exists $e'$ such that $e' <^{\mu} v$ and $e' <^{\mu} e$ then $v \consistency e$.
\label{def:consistency_properties}
\end{definition}
Intuitively,  the child relation equips a hypergraph with a hierarchical structure: certain edges,  labelled with $\bot$,  are meant to interpret the dashed e-boxes of string diagrams.  
These edges can be designated as "parents" to sub-e-hypergraphs,  which interpret the various components of the e-boxes.  
These components are distinguished by the consistency relation, which partitions the maximal sub-e-hypergraph below the hierarchical edge.
% Technically,  condition $(4)$ on the child relations says that $<^{\mu}$ respects connectivity of edges and vertices,  and similarly for the consistency relation. 
An example of an e-hypergraph can be seen below in Figure~\ref{fig:e-cospan-example} (within a grey shaded region).
The e-hypergraph is given by $E = \{e_1, e_2\}, V = \{v_1, \ldots, w_4 \}$, $s$,$t$, $l = \{e_1 \mapsto \bot, e_2 \mapsto f\}$ and
\[
\begin{array}{ccc}
    v_3 <^{\mu} e_1 & v_3 \consistency v_4 & v_3 \not \consistency v_5\\
    \ldots & \ldots & \ldots\\
    w_4 <^{\mu} e_1 & w_3 \consistency w_4 & v_4 \not \consistency w_4
\end{array}
\]
A hierarchical edge is depicted with a dashed box.
The partition induced by $\consistency_{e_1}$ is depicted with a dashed vertical line.
The condition $\consistency_{p} \not = (E_{p} + V_{p}) \times (E_{p} + V_{p})$, in particular, does not allow hierarchical edges with no delimiting lines.
$e_1$ is the maximal edge and hence labelled.
\begin{definition}[E-hypergraph homomorphism]
An \textit{e-hypergraph homomorphism} $\phi: \mathcal{F} \to \mathcal{G}$ between e-hypergraphs $\mathcal{F},\mathcal{G}$ is a pair of functions $\phi_V : V_{\mathcal{F}} \to V_{\mathcal{G}}, \phi_E : E_{\mathcal{F}} \to E_{\mathcal{G}}$ such that
\begin{enumerate}
    \item $\phi$ is a hypergraph homomorphism
    \item for $v \in V_{\mathcal{F}}$ and $e \in E_{\mathcal{F}}$
	    \[
        \text{if } e <_{\mathcal{F}}^{\mu} v \text{ then } \phi(e) <_{\mathcal{G}}^{\mu} \phi(v)
        \]
        and for $e_1, e_2 \in E_{\mathcal{F}}$
        \[
        \text{if } e_1 <_{\mathcal{F}}^{\mu} e_2 \text{ then } \phi(e_1) <_{\mathcal{G}}^{\mu} \phi(e_2)
        \]
        \item for all $x_1, x_2 \in \mathcal{F}$,  $x_1 ~\consistency_{\mathcal{F}}~ x_2$ implies $\phi(x_1) ~\consistency_{\mathcal{G}}~ \phi(x_2)$. 
\end{enumerate}
% If we further have that if $[x_1) = \varnothing$ then $[\phi(x_1)) = \varnothing$ we call the homomorphism \textit{strict}.
\end{definition}
The conditions on e-hypergraph homomorphisms require to preserve immediate parents and the consistency relation.  
\begin{definition}[Category of e-hypergraphs]
The \emph{category of e-hypergraphs},  denoted  $\catname{EHyp(\Sigma)}$,  has e-hypergraphs as objects and e-hypergraph homomorphisms as morphisms.  
\end{definition}

The category of e-hypergraphs has coproducts,  given by the disjoint union of e-hypergraphs,  and an initial object given by the empty e-hypergraph.  
A concrete description of the pushout of two morphisms in this category is given in the Appendix section~\ref{sec:appendix:pushout}.
Generally,  the pushout of two e-hypergraph homomorphisms need not exist,  but we prove that it does in all scenarios that we are interested in, most importantly the pushout for the composition of two cospans with discrete feet exists.

@@

\end{definition}
We will call $n$  \textit{external input interfaces} and $n'$ \textit{internal input interfaces}.
We call $n' \setminus f_{ext}(n)$  the \textit{strictly internal input interfaces}.  
We do analogously for the \emph{output interfaces},  with respect to $m$,  $m'$ and $m' \setminus g_{ext}(m)$.  
We occasionally conflate $f_{ext}$ with $f_{ext};f_{int}$ when it is clear from context,  and also conflate $n$ and $m$ with their images in $n'$ and $m'$,  and likewise $m, m, n', m'$ with their images in $\mathcal{G}$. 
Given an edge $e \in E_\mathcal{G}$ such that $l(e) = \bot$,  we call the \textit{inputs of $e$} the intersection of the strictly internal input interface of $\mathcal{G}$ with the immediate successors of $e$,  and analogously for the \emph{outputs of $e$}.

Consider a relation $R = \{x R y \text{ if } f_{int}(x) \consistency f_{int}(y) \text{ for } x,y \in n'\}$ and let $S$ be its reflexive closure.
The latter partitions $n'$ into non-empty subsets $\{p_{j}\}_{j=1}^{k}$. 
We get an analogous partition for $m'$.
An example of an extended cospan can be seen in Figure~\ref{fig:e-cospan-example}.
The partition of the interfaces is given with colours.
The labels suggest how $f_{int}, f_{ext}$ (respectively, $g_{int}, g_{ext}$) act on vertices.
The ordering of vertices in the interfaces is from left to right.

@@

\begin{definition}
\label{def:iso}
Two extended cospans are isomorphic if there exist isomorphisms $\alpha$, $\beta$ and $\gamma$ making the following diagram commute and such that $\alpha$ and $\gamma$ preserve order within each $p_i$.
\[
\scalebox{0.75}{
    \tikzfig{../figures/combinatorial_semantics/isomorphic_e_cospans}
}
\]
\end{definition}
Intuitively, such construction suggests that the position of the image of $n$ within $n'$ relative to the strict internal interface does not matter; and that when considering individual components of hierarchical edges with their respective portions of strict internal input and output interfaces as plain cospans of hypergraphs (by only considering top-level vertices and edges of these components) they should be isomorphic cospans of hypergraphs in the sense from~\cite{bonchi_string_2022-1}.
Examples of isomorphic and non-isomorphic extended cospans can be found in Appendix section~\ref{sec:appendix:iso}.

@@

Composition of two morphisms 
\begin{align*}
	n \xrightarrow{f_{ext}} n' \xrightarrow{f_{int}} &\mathcal{F} \xleftarrow{f'_{int}} m' \xleftarrow{f'_{ext}} m\\
	m \xrightarrow{g_{ext}} m'\!' \xrightarrow{g_{int}} &\mathcal{G} \xleftarrow{g'_{int}} k' \xleftarrow{g'_{ext}} k
\end{align*} is computed in two stages.
First, $\mathcal{H}$ is computed as the result of the pushout square shown below; 
% https://q.uiver.app/#q=WzAsMTAsWzAsMCwiZXh0KFgpIl0sWzEsMSwiWCJdLFsyLDIsIkEiXSxbMywwLCJZIl0sWzQsMCwiZXh0KFkpIl0sWzUsMCwiWSciXSxbNiwyLCJCIl0sWzcsMSwiWiJdLFs4LDAsImV4dChaKSJdLFs0LDMsIl57XFx1bGNvcm5lcn1BK197Zl97ZXh0fTtmLGdfe2V4dH07Z31CIl0sWzAsMV0sWzEsMl0sWzMsMiwiZiJdLFs0LDMsImZfe2V4dH0iXSxbNCw1LCJnX3tleHR9IiwyXSxbNSw2LCJnIiwyXSxbNyw2XSxbOCw3XSxbMiw5XSxbNiw5XV0=
\[
\trimbox{0cm 0cm 0cm 0.75cm}{
\adjustbox{scale=0.9,center}
{\tikzfig{../figures/combinatorial_semantics/pushout_e_cospans}}
}
\]
then,  the result of composition is defined as follows:
\[
n \xrightarrow{f_{ext};\iota_1} n' + (m'' \setminus m) \xrightarrow{h_{1}} \mathcal{H} \xleftarrow{h_2} k' + (m' \setminus m) \xleftarrow{g'_{ext};\iota_1} k
\]
where $h_i$ are defined as follows,  using $|$ to denote the restriction of a function on a discrete e-hypergraph. 
\[
    h_1 = [ f_{int};p_1, ~(g_{int};p_2)|_{m'' \setminus m} ]
\qquad
    h_2 = [ g'_{int};p_2, ~(f'_{int};p_1)|_{m' \setminus m} ]
\]
% \[
% \begin{tikzcd}
% Y_{ext} \arrow[d] & 0 \arrow[l] \arrow[d]    \\
% Y'          & Y' \setminus Y_{ext} \arrow[l]
% \end{tikzcd}
% \]
% and then the disjoint union is just a coproduct. 
% where $g'$ is the restriction of $g$ to ${Y' \setminus Y_{ext}}$ and $f'$ the restriction of $f$ to $Y \setminus Y_{ext}$. 
The identity of composition is given by the obvious extended cospan of identities. 
$\Ecospans$ inherits a symmetric monoidal structure from the coproduct (and initial object) structure of $\catname{EHyp({\Sigma}})$,  analogously to Definition~\ref{def:cspd}.

@@

\begin{definition}[Category of well-typed MDA e-hypergraphs with interfaces]
We call an extended cospan in $\Ecospans$,  as below,   \textit{monogamous directed acyclic (MDA)} if
\[
n \xrightarrow{f_{ext}} n' \xrightarrow{f_{int}} \mathcal{G} \xleftarrow{g_{int}} m' \xleftarrow{g_{ext}} m
\]
\begin{enumerate}
        \item underlying hypergraph of $\mathcal{G}$ is directed acyclic; 
        \item $f_{int}$, $g_{int}$ are monomorphisms; 
        \item in-degree and out-degree of every vertex is at most 1; 
        \item vertices with in-degree $0$ are precisely the image of $f_{int}$; 
        \item vertices with out-degree $0$ are precisely the image of $g_{int}$; 
        % the item below is not assumed by the definition of a cospan of e-hypergraphs
        % \item $[v) = \varnothing$ for all $v$ in the image of $f_{ext};f$, and for all $v$ in the image of $g_{ext};g$. 
\end{enumerate}
\label{def:monogamy_ehyp}

Further,  consider the sets of input and output vertices,  $I$ and $O$ respectively,  of a hierarchical edge $e$ of an e-hypergraph $\mathcal{G}$. 
The consistency relation of $\mathcal{G}$ partitions $I$ and $O$.  
We call $e$ \emph{well-typed} if the size of each subset in the partition of $I$ is $|s(e)|$ and the size of each partition of $O$ is $|t(e)|$.
We will call an MDA extended-cospan in $\MdaEcospans$ \emph{well-typed} if all hierarchical edges in its carrier are well-typed.
We denote by $\MdaEcospans$ the subcategory of $\Ecospans$ consisting of well-typed MDA extended-cospans. 
\end{definition}
Note,  $\MdaEcospans$ is indeed a category,  and in fact inherits the symmetric monoidal structure of $\Ecospans$:  it is easy to check that identities,  the monoidal unit,  symmetry,  and the composition and tensor product of any two well-typed cospans are all (again) well-typed.

Unlike in the previous section,  in order to give an equivalence with $\catname{S}^+(\Sigma)$,  we must first develop the theory of DPOI rewriting for e-hypergraphs.
This is because the equations for semilattice enrichment are not intended to be subsumed by the e-hypergraph representation, apart from the commutativity equation which is absorbed by the definition of isomorphic cospans.
Instead,  we will treat them via rewriting,  as we do with equations of the signature.  This is because the semilattice equations involve sharing and copying, operations on string diagrams which are not usually quotiented away.
% , cf. string diagrams for Cartesian categories.  
An ideal combinatorial representation would factor out the two semilattice equations which \textit{do not} involve sharing and copying (namely,  associativity and idempotence) --- however, a limitation of the current work is that we do not achieve this.
Presently,  we will treat these equations via rewriting as well.

### 2025-01-21 LICS/sections/introduction.tex
wip

Given a category $\mathbb{C}$  with objects $A,B \in \mathbb{C}$ we denote by $\mathbb{C}(A,B)$ the corresponding hom-set.  
We write the identity morphism on $A$ as $\id_A$,  or simply $A$.  
We commonly write $f;g$ for composition in diagrammatic order.  
Composition in the usual order is written $g \circ f$.  
We denote the tensor product of an SMC $\mathbb{C}$ by $\otimes$,  its unit by $I$, and its symmetry natural transformation as $\sym$ \cite{maclane}.  
We adopt the convention that $\otimes$ binds more tightly than $(;\!)$.  
We elide all associativity and unit isomorphisms associated with monoidal categories,  and often omit subscripts on identities and natural transformations where it can be inferred.  
We denote by $\catname{SLat}$ the closed monoidal category of semilattices which is defined in section~\ref{sec:appendix:1} of the Appendix.

@@

\begin{proposition}(A specialized case of Proposition 6.4.3~\cite{Borceux_1994}. See also Appendix section~\ref{sec:appendix:1})
There is a 2-adjunction 
% https://q.uiver.app/#q=WzAsMixbMCwwLCJcXGNhdG5hbWV7U0xhdHR9LVxcY2F0bmFtZXtDYXR9Il0sWzIsMCwiXFxjYXRuYW1le0NhdH0iXSxbMSwwLCJGIiwyLHsiY3VydmUiOjR9XSxbMCwxLCJHIiwyLHsiY3VydmUiOjR9XSxbMiwzLCIiLDAseyJsZXZlbCI6MSwic3R5bGUiOnsibmFtZSI6ImFkanVuY3Rpb24ifX1dXQ==
\[\begin{tikzcd}
	\catname{SLatt}\text{--}\catname{Cat}\arrow[rr, "\mathcal{U}", bend left] & \hspace{-1em}\top & \catname{Cat} \arrow[ll, "\mathcal{F}", bend left]
	\end{tikzcd}
\]
that is induced by a usual free-forgetful adjunction

>>>>

Given a category $\mathbb{C}$  with objects $A,B \in \mathbb{C}$ we denote by $\mathbb{C}(A,B)$ the corresponding hom-set.  
We write the identity morphism on $A$ as $\id_A$,  or simply $A$.  
We commonly write $f;g$ for composition in diagrammatic order.  
Composition in the usual order is written $g \circ f$.  
We denote the tensor product of an SMC $\mathbb{C}$ by $\otimes$,  its unit by $I$, and its symmetry natural transformation as $\sym$ \cite{maclane}.  
We adopt the convention that $\otimes$ binds more tightly than $(;\!)$.  
We elide all associativity and unit isomorphisms associated with monoidal categories,  and often omit subscripts on identities and natural transformations where it can be inferred.  
We denote by $\catname{SLat}$ the closed monoidal category of semilattices which is defined in section~\ref{sec:appendix:slat} of the Appendix.

@@

\begin{proposition}(A specialized case of Proposition 6.4.3~\cite{Borceux_1994}. See also Appendix section~\ref{sec:appendix:slat})
There is a 2-adjunction 
% https://q.uiver.app/#q=WzAsMixbMCwwLCJcXGNhdG5hbWV7U0xhdHR9LVxcY2F0bmFtZXtDYXR9Il0sWzIsMCwiXFxjYXRuYW1le0NhdH0iXSxbMSwwLCJGIiwyLHsiY3VydmUiOjR9XSxbMCwxLCJHIiwyLHsiY3VydmUiOjR9XSxbMiwzLCIiLDAseyJsZXZlbCI6MSwic3R5bGUiOnsibmFtZSI6ImFkanVuY3Rpb24ifX1dXQ==
\[\begin{tikzcd}
	\catname{SLatt}\text{--}\catname{Cat}\arrow[rr, "\mathcal{U}", bend left] & \hspace{-1em}\top & \catname{Cat} \arrow[ll, "\mathcal{F}", bend left]
	\end{tikzcd}
\]
that is induced by a usual free-forgetful adjunction
